Regional Base-Flow Index in Arid Landscapes Using Machine Learning and Instrumented Records

### 2025-01-29 MANUSCRIPT.qmd

$K$-fold cross-validation provides an unbiased estimate of a model's accuracy on unseen data, while also insuring against overfitting or underfitting. In this approach, the data is randomly divided into $k$ folds of equal size. The model is trained on $k-1$ folds and tested on the remaining fold, referred to as the validation set. This process is repeated $k$ times, with each fold serving as the validation set exactly once. Each iteration trains an independent model with the same hyperparameters but using a different subset of training data. By averaging the model performance across all $k$ folds, we achieve a robust and reliable estimate of predictive accuracy. Root mean squared error (RMSE) was used as the performance metric for both model optimization and evaluation. RMSE provided a consistent and interpretable measure of the model’s accuracy throughout the training and validation process.

![Schematic of $k$-fold cross validation. The dataset is randomly divided into $k$ stratified folds. Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development. Performance metrics for the test set are calculated and recorded, and this process is repeated for all $k$ folds.](images/k-fold_fig.png){#fig-k-fold width="6000"}

@@

The regionalized (HUC-8) long-term BFI (1991–2020) is shown in @fig-bfi-huc, while the long-term BFI for stream reaches with a Strahler stream order of 3 or greater is presented in @fig-bfi-streams. Basins with high BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction. Elevated BFI values are also observed along portions of the Mogollon Rim, a heavily forested region with high precipitation that marks the transition between physiographic regions. Additionally, headwater regions of perennial rivers tend to exhibit higher BFI values. In contrast, low BFI values are found in areas like the Defiance Uplift in northeastern Arizona and the arid southern regions of the state.

![Predicted long-term BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20241203.png){#fig-bfi-huc}

@@

The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consistently low BFI values (\< 0.2). This is likely due to low-yielding perched aquifers underlying the Defiance Uplift in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02]. A notable tendency emerges along most major rivers in the study area: upstream reaches tend to exhibit higher BFI values, while downstream reaches display lower values. This pattern is presumed to result from greater groundwater-surface water interactions at stream headwaters, influenced by spring outlets, and the dilution of base flow as water moves downstream. This trend is particularly evident along the Gila River, Verde River, and Little Colorado River in Arizona.

@@

This study provides new insights into base-flow dynamics and groundwater contributions in Arizona’s dryland rivers by combining an analysis of instrumented streamflow records with machine learning predictions for ungauged basins. The results highlight significant spatial variability in BFI, with approximately 32% of Arizona’s long-term streamflow originating from groundwater discharge. Regions such as the Grand Canyon and Mogollon Rim demonstrate high BFI values due to strong groundwater-surface water interactions, while areas like the Little Colorado River Basin exhibit low BFI values, reflecting limited groundwater recharge.

Using an XGBoost machine learning algorithm, we successfully predicted long-term BFI in ungauged basins, achieving strong model performance (R² = 0.764, RMSE = 0.129). The model performed well across all classifications, demonstrating its robustness in capturing base-flow dynamics across a region with substantial variability in climate, elevation, and physiographic characteristics. Key predictors included elevation, land cover, and soil type, highlighting the importance of integrating hydroclimate and physiographic characteristics into regional hydrological models. These predictions address the limitations posed by Arizona’s sparse streamgage network, offering a scalable approach to estimate BFI in data-limited regions.

Our analysis of BFI trends in gauged catchments revealed that precipitation is the primary driver of base-flow variability, with evapotranspiration and temperature contributing additional complexity. These findings emphasize the critical role of climate-hydrology interactions in shaping groundwater contributions to streamflow. Inverse trends between temperature and BFI suggest that further warming could reduce groundwater contributions to streamflow. Coincident trends in precipitation and BFI further underscore the importance of understanding recharge processes, especially in arid and semi-arid landscapes, where precipitation events play key roles in replenishing groundwater.

>>>>

$K$-fold cross-validation provides an unbiased estimate of a model's accuracy on unseen data, while also insuring against overfitting or underfitting. In this approach, the data are randomly divided into $k$ folds of equal size. The model is trained on $k-1$ folds and tested on the remaining fold, referred to as the validation set. This process is repeated $k$ times, with each fold serving as the validation set exactly once. Each iteration trains an independent model with the same hyperparameters but using a different subset of training data. By averaging the model performance across all $k$ folds, we achieve a robust and reliable estimate of predictive accuracy. Root mean squared error (RMSE) was used as the performance metric for both model optimization and evaluation. RMSE provided a consistent and interpretable measure of the model’s accuracy throughout the training and validation process.

![Schematic of $k$-fold cross validation. The dataset is randomly divided into $k$ stratified folds. Each fold serves as the validation set once, while the remaining folds are combined to create a training set for model development. Performance metrics for the test set are calculated and recorded, and this process is repeated for all $k$ folds.](images/k-fold_fig.png){#fig-k-fold fig-align="center" width="100%"}

@@

The regionalized (HUC-8) long-term BFI (1991–2020) is shown in @fig-bfi-huc, while the long-term BFI for stream reaches with a Strahler stream order of 3 or greater is presented in @fig-bfi-streams. Basins with high BFIs, such as those along the Grand Canyon in the northwestern part of the study area, indicate greater surface water and groundwater interaction. Elevated BFI values are also observed along portions of the Mogollon Rim, a heavily forested region with high precipitation that marks the transition between physiographic regions. Additionally, headwater regions of perennial rivers tend to exhibit higher BFI values. In contrast, low BFI values are found in areas like the Defiance Plateau in northeastern Arizona and the arid southern regions of the state.

![Predicted long-term BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20250129.png){#fig-bfi-huc}

@@

The stream reaches of the Little Colorado River Basin (northeastern Arizona) indicate consistently low BFI values (\< 0.2). This is likely due to low-yielding perched aquifers underlying the Defiance Plateau in northeastern Arizona, which are hydrologically connected to surface streams, while the high-yield, confined regional aquifer is much deeper [@blanchard02]. A notable tendency emerges along most major rivers in the study area: upstream reaches tend to exhibit higher BFI values, while downstream reaches display lower values. This pattern is presumed to result from greater groundwater-surface water interactions at stream headwaters, influenced by spring outlets, and the dilution of base flow as water moves downstream. This trend is particularly evident along the Gila River, Verde River, and Little Colorado River in Arizona.

@@

This study provides new insights into base-flow dynamics and groundwater contributions in Arizona’s dryland rivers by combining an analysis of instrumented streamflow records with machine learning predictions for ungauged basins. The results highlight significant spatial variability in BFI, with approximately 32% of Arizona’s long-term streamflow originating from groundwater discharge. Regions such as the Grand Canyon and Mogollon Rim demonstrate high BFI values due to strong groundwater-surface water interactions, while areas like the Little Colorado River Basin exhibit low BFI values, reflecting limited groundwater recharge (@fig-bfi-huc).

Using an XGBoost machine learning algorithm, we successfully predicted long-term BFI in ungauged basins, achieving strong model performance (R² = 0.764, RMSE = 0.129). The model performed well across all classifications, demonstrating its robustness in capturing base-flow dynamics across a region with substantial variability in climate, elevation, and physiographic characteristics (@tbl-performance). Key predictors included elevation, land cover, and soil type, highlighting the importance of integrating hydroclimate and physiographic characteristics into regional hydrological models (@fig-shap_values). These predictions address the limitations posed by Arizona’s sparse streamgage network, offering a scalable approach to estimate BFI in data-limited regions.

Our analysis of BFI trends in gauged catchments revealed that precipitation is the primary driver of base-flow variability, with evapotranspiration and temperature contributing additional complexity (@tbl-coincident-trends). These findings emphasize the critical role of climate-hydrology interactions in shaping groundwater contributions to streamflow. Inverse trends between temperature and BFI suggest that further warming could reduce groundwater contributions to streamflow. Coincident trends in precipitation and BFI further underscore the importance of understanding recharge processes, especially in arid and semi-arid landscapes, where precipitation events play key roles in replenishing groundwater.
